Gap formation and stability in non-isothermal protoplanetary discs

### 2015-03-26 paper/model.tex
revision 2 done

The total potential $\Phi$ includes the stellar potential, planet potential
(described below) 
and indirect potentials to account for the non-inertial reference
frame. 
In the momentum equations, $\bm{f}_\nu$ represent viscous forces, 
which includes artificial bulk viscosity to handle shocks, and a
Navier-Stokes viscosity whose magnitude is  
characterised by a constant kinematic viscosity parameter
$\nu$. However, we will be considering effectively inviscid discs by
adopting small values of $\nu$.

@@

Note that the energy source terms
have been chosen to be absent at $t=0$, allowing the disc to be
initialised close to steady state. The $\mathcal{C}$ function attempts
to restore the initial energy density (and 
therefore temperature) profile. In practice, this is a cooling term at
the gap edge because disc-planet interaction leads to heating.

@@

\subsubsection{Fourier modes} 
The RWI is characterised by exponentially
growing perturbations. Though in this paper we do not consider a
formal linear instability calculation, modal analysis will be useful
to analyse the growth of perturbations with different azimuthal
wavenumbers, which is associated with the number of vortices initially
formed by the RWI.

>>>>

The total potential $\Phi$ includes the stellar potential, planet potential
(described below) 
and indirect potentials to account for the non-inertial reference
frame. 
In the momentum equations, $\bm{f}_\nu$ represent viscous forces, 
which includes artificial bulk viscosity to handle shocks, and a
Navier-Stokes viscosity whose magnitude is  
characterized by a constant kinematic viscosity parameter
$\nu$. However, we will be considering effectively inviscid discs by
adopting small values of $\nu$.

@@

Note that the energy source terms
have been chosen to be absent at $t=0$, allowing the disc to be
initialized close to steady state. The $\mathcal{C}$ function attempts
to restore the initial energy density (and 
therefore temperature) profile. In practice, this is a cooling term at
the gap edge because disc-planet interaction leads to heating.

@@

\subsubsection{Fourier modes} 
The RWI is characterized by exponentially
growing perturbations. Though in this paper we do not consider a
formal linear instability calculation, modal analysis will be useful
to analyse the growth of perturbations with different azimuthal
wavenumbers, which is associated with the number of vortices initially
formed by the RWI.

### 2015-03-26 paper/results.tex
revision 2 done

\subsection{Non-axisymmetric instability}\label{linear}
We now examine the evolution of the gap for $t>30P_0$, with the
planet potential switched off, but with an added surface density
perturbation. For all three cooling times $\tilde{\beta}=0.1,\, 1.0,\,
10$, we observe exponential growth of non-axisymmetric
structures. An example is shown in Fig. \ref{linearmodes} for 
$\tilde{\beta}=10$. We characterise these
modes with an azimuthal wavenumber $m$ and growth rate $\sigma(m)$ as defined by
Eq.~\ref{fouriertransform}---\ref{growth}. Mode amplitudes were
averaged over $r-r_p\in[2,5]r_h$. Table \ref{modetable}
lists the growth rates measured during 
linear growth for 5 values of $m$ centred around that with maximum
growth rate.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
  0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimensionalized by the initial axisymmetric component 
    $\Sigma_0(t=0)$, for the
    `planet-off' simulation with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth with a growth rate of 
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
  % {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

@@

We characterise the smoothness of the outer gap edge with a dimensionless
gap edge gradient parameter
\begin{align}
  \delta\Sigma(t)= \la \frac{\partial \langle \Sigma(t,r)
    \rangle_{\phi}}{\partial r} \cdot \frac{r}{\langle \Sigma(t=0,r)
    \rangle_{\phi}} \ra_{\Delta r}  
\end{align}
where $\Delta r = r\in [r_p,r_p + 6r_h]$ is the radial range of averaging, spanning
from centre of the gap to the radius of the surface density maximum.
A larger $\delta \Sigma$ characterizes a sharper gap edge and  
larger local surface density maxima.

@@

{\bf We comment here that a locally isothermal simulation, where the
  disc sound-speed is kept constant in time, was also
  performed for comparison. In this case we did not observe
  significant vortex decay within the simulation timescale
 (implying $t_\mathrm{diss}\gtrsim2000P_0$).  We expect corresponding vortex
 lifetime to exceed that for $\tilde{\beta}=0.01$. Including an energy
 equation, even with rapid cooling (or setting $\gamma$ close to
 unity), could lead to discrepencies with a locally isothermal disk due to the
 advection-creation of specific entropy within the planet's horseshoe
 region in the former case,  thereby affecting the generalized vortensity 
 and therefore the vortex evolution. Nevertheless, a longer vortex
 lifetime in a locally isothermal disk would 
 be consistent with the above trend of increasing vortex lifetimes as
 $\tilde{\beta}\to0$. }

@@

%This effect can be seen in Mach number around the vortices where the
%growth of Mach number with respect to time is increasing with cooling rate.
%Thus the intially weaker formed vortices for higher cooling rate grow
%substantialy faster than the sound speed and can shock the system in earlier
%times.
%While our `planet-off' simulations show 
%This may be attributed to the higher temperature
%reached in the latter case 
%
%($h\simeq 0.07$) in comparison with the former
%($h\simeq 0.06$), since \cite{li00} showed that the RWI is favoured by
%larger disc temperature.

>>>>

\subsection{Non-axisymmetric instability}\label{linear}
We now examine the evolution of the gap for $t>30P_0$, with the
planet potential switched off, but with an added surface density
perturbation. For all three cooling times $\tilde{\beta}=0.1,\, 1.0,\,
10$, we observe exponential growth of non-axisymmetric
structures. An example is shown in Fig. \ref{linearmodes} for 
$\tilde{\beta}=10$. We characterize these
modes with an azimuthal wavenumber $m$ and growth rate $\sigma(m)$ as defined by
Eq.~\ref{fouriertransform}---\ref{growth}. Mode amplitudes were
averaged over $r-r_p\in[2,5]r_h$. Table \ref{modetable}
lists the growth rates measured during 
linear growth for 5 values of $m$ centred around that with maximum
growth rate.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
  0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimensioqnalized by the initial axisymmetric component 
    $\Sigma_0(t=0)$, for the
    `planet-off' simulation with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth with a growth rate of 
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
  % {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

@@

We characterize the smoothness of the outer gap edge with a dimensionless
gap edge gradient parameter
\begin{align}
  \delta\Sigma(t)= \la \frac{\partial \langle \Sigma(t,r)
    \rangle_{\phi}}{\partial r} \cdot \frac{r}{\langle \Sigma(t=0,r)
    \rangle_{\phi}} \ra_{\Delta r}  
\end{align}
where $\Delta r = r\in [r_p,r_p + 6r_h]$ is the radial range of averaging, spanning
from centre of the gap to the radius of the surface density maximum.
A larger $\delta \Sigma$ characterizes a sharper gap edge and  
larger local surface density maxima.

@@

{\bf We comment here that a locally isothermal simulation, where the
  disc sound-speed is kept constant in time, was also
  performed for comparison. In this case we did not observe
  significant vortex decay within the simulation timescale
 (implying $t_\mathrm{diss}\gtrsim2000P_0$).  We expect the corresponding vortex
 lifetime to exceed that for $\tilde{\beta}=0.01$. Including an energy
 equation, even with rapid cooling (or setting $\gamma$ close to 
 unity), could still lead to discrepancies with a locally isothermal disc. This is 
 due to the advection-creation of specific entropy within the planet's horseshoe
 region with the former case,  thereby affecting the generalized vortensity 
 and therefore the instability (\S\ref{linear}) and subsequent vortex evolution. 
 Nevertheless, a longer vortex 
 lifetime at the gap edge in a locally isothermal disk would 
 be consistent with the above trend of increasing vortex lifetimes as
 $\tilde{\beta}\to0$.}

@@

%This effect can be seen in Mach number around the vortices where the
%growth of Mach number with respect to time is increasing with cooling rate.
%Thus the initially weaker formed vortices for higher cooling rate grow
%substantialy faster than the sound speed and can shock the system in earlier
%times.
%While our `planet-off' simulations show 
%This may be attributed to the higher temperature
%reached in the latter case 
%
%($h\simeq 0.07$) in comparison with the former
%($h\simeq 0.06$), since \cite{li00} showed that the RWI is favoured by
%larger disc temperature.
